Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-04 paper/paper.tex
revision

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added to convert itself to type @State@.

Function @step@ now becomes of type @NDState => Option[NDState]@. Returning a @None@ value means
reaching the halt.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all the closures at this fork point share the same stack structure and return point.

@@

If the continuation @konts@ is empty, then it means we have reached an end of the computation;
otherwise, we should return the values of @ae@ to its caller which is contained in the top
frame of the stack. However, since a meta-continuation @ndk@ is introduced, we need to dispatch
more cases when @konts@ is empty:

\begin{itemize}
  \item If @ndk@ is also empty, then we have reached the end of all computations. Besides,
    the control component @ae@ now is an atomic value, therefore @None@ is returned.
  \item If the target closure set of the front-most @NDCont@ is empty (@Nil@ in the code),
    then we have computed all the closures at this fork point and should move to the next fork 
    point. Since the meta-continuation @nkd@ is represented by a list, and we always
    append new elements to its front, we resume to the most recent fork point by
    popping up the front-most element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, then build a new environment and store for 
    the body expression of that closure. We use the same frame list and time which are copied 
    from the fork point rather than current one.
    The meta-continuation @new_ndk@ is updated by the removal of that popped closure, which makes
    sure the closure set will eventually shrink to empty.
\end{itemize}

@@

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ takes an @NDState@ and a set of explored states @seen@ 
as arguments and returns a set of reachable states once it terminates.
It is is essentially formed by merging the functionality of @step@ into the @drive@ function.
Since each case in function @step@ are statically known, we inline the pattern match on the
@Option@ type and replace it with @new_seen@ or a recursive call to @drive_step@ for @None@
or @Some@, respectively.

@@

Since there are two layers of continuations, we disentangle them into three mutually recursive 
functions:
@drive_step@, which plays the same role as before; 
@continue@, which is invoked from @drive_step@ when encountering an atomic expression, 
  and dispatches the reduction context of the analyzed language; and 
@ndcontinue@, which is invoked from @continue@ when the normal stack is empty,
  and dispatches the meta-continuations @NDCont@.

@@

Function @ndcontinue@ is identical to the dispatching logic in function @continue@ when @konts@ is empty.
When @ndk@ is empty, the current explored states @seen@ is returned; the analysis then terminates.
\iffalse
The shape for @ndcontinue@ is similar to @continue@; after all, they both use a 
@List@ representation for continuations. When the closure set is empty, we recursively 
call @drive_step@ with the rest of @ndk@, since in this case the continuation 
@konts@ is still empty, so the state will be dispatched to @ndcontinue@ again.
Otherwise, we invoke @drive_step@ with the a new @NDState@ where the control 
is transferred to the body expression of the next closure object.
\fi

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined 
language be part of the control flows of the meta-language.

@@

First of all, as we are getting close to achieving abstract definitional interpreters, 
the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).
Since there will be no first-order data types representing model of stacks,
we introduce a configuration class @Config@ as a state-like definition by eliminating
@konts@ and @ndk@ from the definition of state. The type of return value of the abstract
interpreter accordingly becomes @Set[Config]@.

@@

By looking at the definition of the first-order function @continue@ from the last section,
we may find it calls the meta-continuation @ndcontinue@ inside of its body. 
This observation leads us to define the types of higher-order continuation and meta-continuation:

>>>>

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added to convert itself to type @State@.  
Function @step@ now becomes of type @NDState => Option[NDState]@. Returning a @None@ value means
reaching the halt.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is constructed,
which contains the rest of closures, the values of evaluating @ae@, the store, the time,
and the new frame list. We use the store before updating, because different 
closures may form different binding addresses @baddr@. We also use the \textit{new} frames,
because all the closures at this fork point share the same stack structure and return point.

@@

If the continuation @konts@ is empty, then the machine has reached an end of the computation;
otherwise, we should return the values of @ae@ to its caller which is contained in the top
frame of the stack. However, since a meta-continuation @ndk@ is introduced, we need to dispatch
more cases when @konts@ is empty:

\begin{itemize}
  \item If the meta-continuation @ndk@ is also empty, then the machine has reached the end of \textit{all} 
    computations. Besides, the control component @ae@ now is an atomic value, therefore @None@ is returned.
  \item If the target closures set of the front-most @NDCont@ is empty (@Nil@ in the code),
    then we have computed all the closures at this fork point and should move to the next fork 
    point. Since the meta-continuation @nkd@ is represented by a list, and we always append new 
    elements to its front, we resume to the most recent fork point by popping up the front-most 
    element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, and build a new environment and store for 
    evaluating the body expression of that closure. We use the same frame list and time which 
    are copied from the fork point rather than current one.
    The meta-continuation @new_ndk@ is updated by the removal of that popped closure, which makes
    sure the closure set will eventually shrink to empty.
\end{itemize}

@@

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ takes an @NDState@ and a set of explored states @seen@ 
as arguments and returns a set of reachable states once it terminates.
It is essentially formed by merging the functionality of @step@ into the @drive@ function.
Since each case in function @step@ are statically known, we can inline the pattern match on the
@Option@ type and replace it with returning @new_seen@ when @None@ is matched, or calling 
@drive_step@ recursively when @Some@ is matched.

@@

Since there are two layers of continuations, we disentangle them into three mutually recursive 
functions:
1) @drive_step@, which plays the same role as before; 
2) @continue@, which is invoked from @drive_step@ when encountering an atomic expression, 
  and dispatches the reduction context of the analyzed language; and 
3) @ndcontinue@, which is invoked from @continue@ when the normal stack is empty,
  and dispatches the meta-continuation @NDCont@.

@@

Function @ndcontinue@ is identical to the dispatching logic in function @continue@ when @konts@ is empty.
When @ndk@ is empty, the current explored states @seen@ is returned; the analysis then terminates.
Otherwise, a new @NDState@ is constructed depending on the status of the closures set.
% first call to drive_step can be replaced with continue. ae is atomic, konts is empty.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined 
language be part of the control flows of the meta-language.
In this section, we first present a direct refunctionalization of the continuations.
Next, we adopt a caching algorithm to obtain a computable analysis\cite{darais2017abstracting},
and further prepare the interpreter for tranforming to direct-style later.

@@

First of all, as we are getting close to achieving abstract definitional interpreters, 
the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).
Since there will be no first-order data types representing model of stacks,
we introduce a configuration class @Config@ as a state-like definition by eliminating
@konts@ and @ndk@ from the definition of state. Accordingly, the abstract interpreter
now returns @Set[Config]@.

@@

By looking at the definition of the first-order function @continue@ from the last section,
we may find it calls the meta-continuation @ndcontinue@ inside of its body.
This observation leads us to define the types of higher-order functions representing 
continuation and meta-continuation:
